Latent Predicate Networks: Concept Learning with Probabilistic Context-Sensitive Grammars

### 2014-10-20 drafts/cfp/draft.tex
Changes to draft.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{anyfontsize}
\usepackage{subcaption}
\usepackage{graphicx}
\title{Latent Predicate Networks: Concept Learning with Probabilistic Context-Sensitive Grammars}
\author{Eyal Dechter \& Josh Rule \& Josh Tenenbaum}

@@

\section{Abstract}
To what extent can the meaning of language be learned from linguistic
data alone?  The sentences that people say contain a great deal of
information about the meaning of linguistic concepts. Most approaches
to semantics assume an explicit knowledge representation formalism
that ties utterances to propositions about things in the world. The
study of linguistic syntax, however, has largely assumed that the
meaning of syntactic structures is encoded directly into the grammar
of a language. For machine learning, extending the grammatical
embedding of meaning to at least some of the linguistic concepts is
attractive because it reduces the problem of automated language
learning to the problem of learning a distribution over naturally
occuring utterances. To what extent is this a feasible approach? Here,
we present a first attempt at the automated induction of probabilistic
context sensitive grammars which we call Latent Predicate Networks
(LPNs), motivated by the idea that the predicates and rules that are
learned correspond to the concepts expressed by utterances. We show
that for a small fragment of sentences expressing relationships
between English number words, we can use hierarchical Bayesian
inference to learn LPNs that can answer simple queries about
previously unseen linguistic concepts within this domain. These
generalizations, we argue, demonstrate LPNs' promise as a tool for
learning and representing conceptual knowledge in language.

% A popular approach in statistical machine learning is to automatically
% find the latent \emph{features} of the data that are most relevant or
% informative. But a latent feature based view of concept learning does
% not match up naturally with the structured symbolic knowledge that
% seems to underlie people's understanding of many domains. How can we
% combine the flexibility, noise tolerance, and computational
% practicality of latent feature based statistical machine learning with
% the symbolic knowledge representations that best capture human
% understanding? Here present a learning model we dub Latent Concept
% Grammars (LCGs) as a proposal. An LCG is a densely connected set of
% probabilistic string rewrite rules over an initially uninterpreted set
% of predicates. When fit to data using a hierarchical Bayesian prior,
% these predicates assume the roles of informative latent concepts. This
% approach allows us to learn knowledge that is stochastic and
% recursively structured while maintaining computational
% tractability. We show how learning in LCGs can be implemented using
% the probabilistic logic programming system PRISM, and we present
% preliminary experiments.

% How do people learn systems of related concepts from
% data?  Consider, for example, a school-age child's knowledge of
% number. This knowledge supports a large number of related operations:
% the child knows which words denote numbers, how to use those words to
% label cardinalities of sets, how to translate between digit
% representations and word representations of number, how to compare
% cardinalities, etc. The child knows how to perform these operations
% over an unbounded set of elements, even though the data that underlies
% this knowledge is small and noisy. Here we propose that such systems
% of related concepts can be learned via hierarchical probabilistic
% inference over probabilistic string rewrite systems (PSRS) -- a mildly
% context-sensitive class of probabilistic string gramamrs that
% corresponds to a restricted fragment of stochastic logic programs
% (SLPs). Specifically, we extend the variational
% expectation-maximization algorithm for Hierarchical Dirichlet PCFGs to
% accomodate PSRSs. This algorithm we present enables a tractible
% compromise between highly restrictive context-free knowledge
% representations and the intractability of performing similar inference
% procedures over general logic programs. Empirically, we show how such
% restricted string rewrite systems can capture non-trivial conceptual
% knowledge and how our algorithm is able to learn such knowledge from
% data.

@@

\section{Introduction}
The vast majority of the models of concept acquisition that have been
proposed in the artificial intelligence, machine learning and
cognitive science literatures cannot represent the conceptual systems
that even young children learn. Take, for example, the school-age
child's knowledge of number. This child knows that there is no largest
number; that ``one million forty three'' and ``8439''denote numbers
but that ``one million forty billion'' and ``a82bd'' do not; that no
two cardinalities refer to the same quantity but that some number
words refer to the same cardinalities (e.g. ``two thousand'' and
``twenty-one hundred''). Moreover, the child knows how to guage the
approximate number of elements in a collection and how to count this
number exactly. In short, the child's knowledge of number consists of
a complex system of concepts over an unbounded set of entities
(i.e. the numbers).

Children are not born with this kind of knowledge, and, since number
is a relatively recent cultural invention, nor is it likely that
children possess innate domain specific background knowledge that
makes it easier to acquire. Children only ever see a tiny fraction of
all the numbers they can reason about and this evidence is biased,
noisy, and can be unreliable.

A fundamental challenge for cognitive science -- and for any field
interested in understanding human-like concept acquisition -- is to
provide a theory of how such complex systems of concepts can be
learned from sparse and noisy data.

People are able to learn about complex systems of related
concepts from small amounts of noisy data. A compelling approach to
understanding such learning is to frame it as Bayesian inference over
probabilistic knowledge bases. Advances in context free grammar
induction in the past couple of decades can be seen as an instance of
this approach, in which

@@

\subsection{Related Work}
Our core inference algorithm borrows heavily from recent algorithms for nonparametric PCFG induction, as well as work extending those algorithms to SLPs~\cite{DBLP:conf/emnlp/LiangPJK07,goldwater2006contextual,johnson2006adaptor,cussens2001parameter}. Like Johnson and Liang, we are interested in inferring the most likely grammar given some collection of sentences, and in doing so without a pre-specified limit on the complexity of the grammar. Like Cussens, we are interested in efficiently extending this algorithm to more powerful and expressive formalisms than CFGs.

@@

An LCG is a hierarchical Bayesian model of string tuples. It is an
extension of the Hierarhical Dirichlet PCFG model (HD-PCFG) to
Probabilistic Range Concatenation Grammars (PRCGs). Whereas the
HD-PCFG is a distribution over strings, an LCG is a distribution over
string tuples belonging to one of a set of observable
relations.

We extend RCGs to Probabilistic Range Concatenation Grammars (PRCGs)
by annotating each clause $C_k \in P$ with probabilities $p_k$ such
that for all predicates $A \in N$ $\sum_{k:head(C_k)=A} p_k = 1$. A
PRCGs defines a distribution over strings $x$ by sampling from
derivations of $S(x)$ according to the product of probabilities of
clauses used in that derivation. This is a well defined distribution
as long as no probability mass is placed on derivations of infinite
length; in this paper, we only consider PRCGs with derivations of
finite length, so we need not worry about this requirement.

@@

\section{Inference}
Inference in hierarchical Dirichlet processes in stochastic grammars
and stochastic logic programs have been well studied. Variational
Expectation-Maximization (VB-EM) is a popular approach, and this is
the algorithm we adopt here. The algorithm is described extensively
elsewhere (CITATION). As we will show in the following section, our
latent stochastic grammar model can be encoded as a stochastic logic
program, so that any VB-EM algorithm for that domain will suffice.

@@

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{anyfontsize}
\usepackage{subcaption}
\usepackage{graphicx}
\title{Latent Predicate Networks: Concept Learning with Probabilistic Context-Sensitive Grammars}
\author{Eyal Dechter,  Josh Rule, Josh Tenenbaum}

@@

\section{Abstract}
How can machines learn about a system of concepts from sentences about
those concepts? One approach is to estimate the distribution from
which those sentences are drawn, and, in doing so, infer the latent
concepts and relationships that best explain the data. We present
early work on a learning framework called Latent Predicate Networks
(LPNs) which attempts to learn concepts by inferring the parameters of
probabilistic context-sensitive grammars that best describe sentences.
We show that for a small fragment of sentences expressing
relationships between English number words, we can use hierarchical
Bayesian inference to learn LPNs that can answer simple queries about
previously unseen linguistic concepts within this domain. These
generalizations, we argue, demonstrate LPNs' promise as a tool for
learning and representing conceptual knowledge in language.

% A popular approach in statistical machine learning is to automatically
% find the latent \emph{features} of the data that are most relevant or
% informative. But a latent feature based view of concept learning does
% not match up naturally with the structured symbolic knowledge that
% seems to underlie people's understanding of many domains. How can we
% combine the flexibility, noise tolerance, and computational
% practicality of latent feature based statistical machine learning with
% the symbolic knowledge representations that best capture human
% understanding? Here present a learning model we dub Latent Concept
% Grammars (LCGs) as a proposal. An LCG is a densely connected set of
% probabilistic string rewrite rules over an initially uninterpreted set
% of predicates. When fit to data using a hierarchical Bayesian prior,
% these predicates assume the roles of informative latent concepts. This
% approach allows us to learn knowledge that is stochastic and
% recursively structured while maintaining computational
% tractability. We show how learning in LCGs can be implemented using
% the probabilistic logic programming system PRISM, and we present
% preliminary experiments.
}
% How do people learn systems of related concepts from
% data?  Consider, for example, a school-age child's knowledge of
% number. This knowledge supports a large number of related operations:
% the child knows which words denote numbers, how to use those words to
% label cardinalities of sets, how to translate between digit
% representations and word representations of number, how to compare
% cardinalities, etc. The child knows how to perform these operations
% over an unbounded set of elements, even though the data that underlies
% this knowledge is small and noisy. Here we propose that such systems
% of related concepts can be learned via hierarchical probabilistic
% inference over probabilistic string rewrite systems (PSRS) -- a mildly
% context-sensitive class of probabilistic string gramamrs that
% corresponds to a restricted fragment of stochastic logic programs
% (SLPs). Specifically, we extend the variational
% expectation-maximization algorithm for Hierarchical Dirichlet PCFGs to
% accomodate PSRSs. This algorithm we present enables a tractible
% compromise between highly restrictive context-free knowledge
% representations and the intractability of performing similar inference
% procedures over general logic programs. Empirically, we show how such
% restricted string rewrite systems can capture non-trivial conceptual
% knowledge and how our algorithm is able to learn such knowledge from
% data.

@@

\section{Introduction}

As Claude Shannon demonstrated with his guessing game experiment in
1950~\cite{Shannon1950}, an adult English speaker's understanding of
her language and her world allows her to reveal that the entropy of a
character in English text is only a little more than one
bit. Plausibly, any machine learning system that were able to predict
natural language text with such accuracy will have learned much about
the concepts expressed in that text. This intuition motivates us to
invert Shannon's experiment and attempt to learn about linguistic
concepts by learning a probabilistic generative model for natural
language.

Probabilistic Context Free Grammars (PCFGs) are popular and much
studied generative models for natural language
syntax~\cite{Someone}. And though there ongoing debate about their
ability to express the syntax of natural language, there is no doubt
that they cannot express dependencies between words that depend upon
their meaning. Consider an example from the domain of sentences
addressed in this paper: ``after twenty eight comes twenty nine.''
The second ``twenty'' depends on the first, and the ``nine'' depends
on the ``eight,'' and this dependency pattern is a clear example of a
\emph{cross-serial dependency}, a dependency pattern that context-free
grammars do not support.

Our approach, then, will be to explore the use of probabilistic
context \emph{sensitive} grammars for modeling natural language,
extending the tools of hierarchical grammar induction~\ref{Cite} from
context-free grammars to ones that can express not only syntactic
relationships but semantic ones as well. In the place of latent
syntactic categories, we will refer to latent predicates -- relations
between sentence components -- that, when learned, should correspond
to the concepts that underlie sentence semantics.

To explore this approach, we introduce a probabilisitic model called a
Latent Predicate Network (LPN), and describe an inference procedure to
learn its parameters using the probabilistic logic programming
language PRISM. Finally, we study its performance when trained on
sentences describing the relationships between numbers.

\section{Latent Predicate Networks}

An LPN is a hierarchical Bayesian model of strings. It is an
extension of the Hierarhical Dirichlet PCFG model (HD-PCFG) to
Probabilistic Range Concatenation Grammars (PRCGs).

We extend RCGs to Probabilistic Range Concatenation Grammars (PRCGs)
by annotating each clause $C_k \in P$ with probabilities $p_k$ such
that for all predicates ${A \in N$ $\sum_{k:head(C_k)=A} p_k = 1}$. A
PRCGs defines a distribution over strings $x$ by sampling from
derivations of $S(x)$ according to the product of probabilities of
clauses used in that derivation. This is a well defined distribution
as long as no probability mass is placed on derivations of infinite
length; in this paper, we only consider PRCGs with derivations of
finite length, so we need not worry about this requirement.

% The vast majority of the models of concept acquisition that have been
% proposed in the artificial intelligence, machine learning and
% cognitive science literatures cannot represent the conceptual systems
% that even young children learn. Take, for example, the school-age
% child's knowledge of number. This child knows that there is no largest
% number; that ``one million forty three'' and ``8439''denote numbers
% but that ``one million forty billion'' and ``a82bd'' do not; that no
% two cardinalities refer to the same quantity but that some number
% words refer to the same cardinalities (e.g. ``two thousand'' and
% ``twenty-one hundred''). Moreover, the child knows how to guage the
% approximate number of elements in a collection and how to count this
% number exactly. In short, the child's knowledge of number consists of
% a complex system of concepts over an unbounded set of entities
% (i.e. the numbers).

% Children are not born with this kind of knowledge, and, since number
% is a relatively recent cultural invention, nor is it likely that
% children possess innate domain specific background knowledge that
% makes it easier to acquire. Children only ever see a tiny fraction of
% all the numbers they can reason about and this evidence is biased,
% noisy, and can be unreliable.

% A fundamental challenge for cognitive science -- and for any field
% interested in understanding human-like concept acquisition -- is to
% provide a theory of how such complex systems of concepts can be
% learned from sparse and noisy data.

% People are able to learn about complex systems of related
% concepts from small amounts of noisy data. A compelling approach to
% understanding such learning is to frame it as Bayesian inference over
% probabilistic knowledge bases. Advances in context free grammar
% induction in the past couple of decades can be seen as an instance of
% this approach, in which

@@

\subsection{Related Work}
Johnson and Liang, we are interested in inferring the most likely
grammar given some collection of sentences, and in doing so without a
pre-specified limit on the complexity of the grammar. Like Cussens, we
are interested in efficiently extending this algorithm to more
powerful and expressive formalisms than CFGs.

@@

@@

\section{Inference}
Inference in hierarchical Dirichlet processes in stochastic grammars
and stochastic logic programs has been an active area of research in
recent
decades~\cite{DBLP:conf/emnlp/LiangPJK07,goldwater2006contextual,johnson2006adaptor,cussens2001parameter}.
Variational inference is a popular approach in this domain and it is
the approach we adopt here. As we explain in the next section, we
implemented inference by translating LPNs into PRISM programs, and we
use the builtin Variational Bayes Expectation-Maximization
algorithm~\cite{sato2008variational}.

@@

\bibliographystyle{amsplain}
\bibliography{draft}

\nocite{*}
